%%%
%%% Annual Cognitive Science Conference
%%% Sample LaTeX Paper -- Proceedings Format
%%%

% Original : Ashwin Ram (ashwin@cc.gatech.edu)       04/01/1994
% Modified : Johanna Moore (jmoore@cs.pitt.edu)      03/17/1995
% Modified : David Noelle (noelle@ucsd.edu)          03/15/1996
% Modified : Pat Langley (langley@cs.stanford.edu)   01/26/1997
% Latex2e corrections by Ramin Charles Nakisa        01/28/1997
% Modified : Tina Eliassi-Rad (eliassi@cs.wisc.edu)  01/31/1998
% Modified : Trisha Yannuzzi (trisha@ircs.upenn.edu) 12/28/1999
% Modified : Mary Ellen Foster (M.E.Foster@ed.ac.uk) 12/11/2000
% Modified : Ken Forbus                              01/23/2004
% Modified : Eli M. Silk (esilk@pitt.edu)            05/24/2005
% Modified : Niels Taatgen (taatgen@cmu.edu)         10/24/2006
% Modified : David Noelle (dnoelle@ucmerced.edu)     11/19/2014
% Modified : Roger Levy (rplevy@mit.edu)             12/31/2018
% Modified : Stephanie Denison                       11/29/2025
% Modified : Dae Houlihan (daeda@mit.edu)            12/01/2025


%%% Change "letterpaper" in the following line to "a4paper" if you must.

\documentclass[10pt,letterpaper]{article}

\usepackage{cogsci}

% \cogscifinalcopy %%% Uncomment this line for the final submission

%%% Bibliography %%%
\usepackage[natbib=true]{biblatex}
\addbibresource{qualia.bib}
\setlength{\bibhang}{.125in}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float} %%% Roger Levy added this and changed figure/table placement to [H] for conformity to Word template, though floating tables and figures to top is still generally recommended!
\usepackage{comment}

% Sometimes it can be useful to turn off hyphenation for purposes such as spell checking of the resulting PDF.
% \usepackage[none]{hyphenat} %%% Uncomment to turn off hyphenation

\title{MaryVLM: Towards a More Rigorous Science of Subjectivity}

%%% Format authors using helper functions from authblk package %%%
\author[1]{\mbox{Kendall Park (kendall@cs.wisc.edu)}}
\author[1]{\mbox{Fred Sala (fred@cs.wisc.edu)}}
\affil[1]{Department of Computer Science, University of Wisconsin-Madison}

%%% Or, format authors manually %%%
% \author{
%   {\large\bfseries Author N. One (a1@uni.edu)$^1$ \& Author Number Two$^2$} \\
%   {\normalsize\normalfont
%     $^1$Department of Hypothetical Sciences, University of Illustrations \\
%     $^2$Department of Example Studies, University of Demonstrations
%   }
% }

\begin{document}

\maketitle

\begin{abstract}
    Include no author information in the initial submission, to facilitate blind review. AI tools cannot
    be listed as authors, and authors retain full responsibility for the accuracy, integrity, and
    originality of all content in their manuscripts. This includes verifying factual claims, ensuring
    proper attribution of ideas, and confirming that the work meets standards for academic integrity and
    does not contain plagiarized content. See the Acknowledgments section of the template for AI use
    declaration and acknowledgment. The abstract should be one paragraph, no more than 150~words,
    indent 1/8~inch on both sides, in 9~point font with single spacing. The heading
    ``\textbf{Abstract}'' should be 10~point, bold, centered, with one line of space below it. This
    one-paragraph abstract section is required only for standard proceedings papers. Following the
    abstract should be a blank line, followed by the header ``\textbf{Keywords:}'' and a list of
    descriptive keywords separated by semicolons, all in 9~point font, as shown below.

    \textbf{Keywords:}
    qualia; consciousness; vision-language models; predictive coding; explanatory gap; Mary's room
\end{abstract}

\input{sections/01_introduction}
\input{sections/02_prior_work}
\input{sections/03_methods}
\input{sections/04_results}
\input{sections/05_discussion}

\section{Acknowledgments}

In the \textbf{initial submission}, please only include acknowledgments of AI use and no other
acknowledgments to preserve anonymity. Regarding AI use: Authors may use AI tools when developing
their projects and preparing their manuscripts, but such use must be described, transparently and
in detail, in either the Methods or Acknowledgments section, as appropriate. Tools that are used to
improve spelling, grammar, and general editing are not included in the scope of these guidelines.
In the \textbf{final submission}, place acknowledgments (including human and AI contributions, and
funding information) in a section \textbf{at the end of the paper}.

\section{References}

\printbibliography

\end{document}
